Ground state energy of quantum dots using the coupled cluster method

### 2018-04-08 project_2/article/article.tex
Initial commit project 2

\title{Variational Monte Carlo on Bosonic Systems}
\author{Winther-Larsen, Sebastian Gregorius}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\author{Schøyen, Øyvind Sigmundson}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\date{\today}

\begin{abstract}
    In this project we use Quantum Variational Monte Carlo on a toy model of a
    bosonic gas  in an elliptic trap. The framework we have built is first
    tested against analytic solutions of simple non-interacting systems. From
    there the system is extended to include perturbed harmonic oscillator traps
    and particle interaction. We compute one-body densities for the interacting
    case and discuss why this tool is important and how our toy model has
    similar attributes to real-world systems.
\end{abstract}

@@

\section{Introduction}
    We will in this project study the Variational Monte Carlo method, and use it
    to evaluate the ground state energy of a trapped, hard sphere Bose gas.

    By building up an increasingly intricate system we eloquently illustrate the
    strength of the Variational Monte Carlo methods. Any Monte Carlo method
    draws on the fact that the computer will not complain when being tasked to
    perform a very similar task up to several million times in rapid succession.
    Monte Carlo methods is in essence a series of guesses where the current
    guess is only a minor alteration of the one prior to it. The trick lies in
    only accepting the subsequent guesses that improves upon our configuration
    up to a certain likelihood.

    The method we have employed is called a \emph{Variational} Monte Carlo
    method, because we make an educated guess at a trial wavefunction which we
    allow certain degrees of freedom called \emph{variational parameters}. We
    shall see that the system converges to a proper \footnote{Global, as far as
    we can tell.} minimum for only a certain set of parameters.

    First, we provide the theoretical foundation for the Variational Monte Carlo
    method. This includes explaining how Monte Carlo integration works, and
    introducing the important quantities \emph{local energy}, \emph{drift force}
    and \emph{one-body densities}.

    Second, we outline in more detailed algebra the nature of the systems we
    will be studying. For the non-interacting system we are able to derive an
    exact variational energy to which we can compare the results from our
    simulations. We furthermore give a description of an interacting system
    implemented as if each particle is a hard sphere.

    Third, we provide a thorough description of algorithms and statistical
    methods employed in this study. Here we introduce the Metropolis-Hastings
    algorithm. This algorithm can by a deft trick be expanded with the method of
    \emph{importance sampling} yielding a more intelligent choice for the next
    step in the Monte Carlo cycling.  The main point of note with regards to the
    statistical analysis is that the data generated by the Monte Carlo cycling
    is autocorrelated and computation of variance must be done with the
    \emph{blocking method} for time series data.  As an automated way of finding
    the optimal variational parameter, we sketch out the method of
    \emph{gradient descent}.

    Fourth, we put the machinery described in the previous sections to work. We
    test against the actual values in a non-interacting system before analyzing
    the more interesting perturbed-trap system with interacting particles.

    Lastly, we highlight interesting aspects of this study in a discussion of
    the results before ending on some summary remarks.

\section{Variational Monte Carlo}
    The variational principle states that for a given \emph{trial wavefunction},
    $\ket{\Psi_T}$, the expectation value of the Hamiltonian, $H$, will be an
    upper bound to the ground state energy, $E_0$ of the system.
    \begin{align}
        E_0 \leq E
        = \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}},
    \end{align}
    The complicated part of the variational principle is finding the trial
    wavefunction that minimizes the energy. The strength in this method lies in
    introducing variational parameters in the trial wavefunction. Using these as
    degrees of freedom we can minimize the energy with respect to these in order
    to find an infimum to the ground state energy for the given trial
    wavefunction and Hamiltonian. For a select few systems we can calculate this
    analytically, but for the interesting systems we quickly hit an intractable
    wall as the systems become analytically unsolvable. To get around this we
    try to calculate the energy for a given Hamiltonian and a trial wavefunction
    for different variational parameters numerically.

    To determine if we have located a minimum we can use the variance of the
    variational energy.
    \begin{align}
        \sigma^2
        &=
        \frac{
            \bra{\Psi_T}H^2\ket{\Psi_T}
        }{
            \bra{\Psi_T}\ket{\Psi_T}
        }
        - \para{
            \frac{
                \bra{\Psi_T}H\ket{\Psi_T}
            }{
                \bra{\Psi_T}\ket{\Psi_T}
            }
        }^2.
    \end{align}
    If we have found the "true" trial wavefunction $\ket{\Psi_T}$ with the
    correct variational parameters, i.e., the eigenfunction to the Hamiltonian
    we get $\sigma^2 = 0$. We are thus interested in trying to find the
    variational parameters in the trial wavefunction that minimizes the
    variance.

    \subsection{Monte Carlo integration}
        Unfortunately the evaluation of the expectation value of the energy for
        a given trial wavefunction involves a large integral.
        \begin{align}
            E
            &= \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
            =
            \frac{\int\dd\vf{r}\Psi_T^{*}(\vf{r})H\Psi_T(\vf{r})}
            {\int\dd\vf{r}|\Psi_T|^2},
        \end{align}
        where $\vf{r} = (\vf{r}_1, \vf{r}_2, \dots \vf{r}_N)$ are the
        positions\footnote{This can include both spatial and spin parts, but we
        will in the remainder of this project ignore spin.} of the particles
        contained in the system.

        To get around\footnote{Not exactly \emph{get around} per se. We still
        have to evaluate the integral, but we make the problem solvable at all.}
        the trouble of evaluating the multidimensional-integral we employ the
        method of Monte Carlo integration. This consists of approximating an
        integral of the expectation value of a certain quantitity by a sum.
        \begin{align}
            \expval{x}
            &=
            \int_{\mathbb{R}}xp(x)\dd x
            \approx
            \frac{1}{N}\sum_{i = 1}^N x_ip(x_i),
        \end{align}
        where $x$ is a random variable with a probability density function
        $p(x)$ and $\{x_i\}$ for $i \in \{1, \dots, N\}$ is the set of drawn
        random variables. Here $N$ is the number of Monte Carlo samples chosen
        to approximate the integral. By the law of large numbers we expect that
        \begin{align}
            \lim_{N \to \infty}\frac{1}{N}\sum_{i = 1}^N x_ip(x_i)
            = \expval{x}.
        \end{align}

        For us to be able to take Monte Carlo integration to the quantum realm
        we have to construct a probability density function from the
        wavefunction. Luckily the wavefunction squared is just such a quantity.
        Given a normalized wavefunction $\ket{\psi}$ we can compute the
        expectation value of a quantum operator $\mathcal{O}$ using Monte
        Carlo integration.
        \begin{align}
            \expval{\mathcal{O}}
            &= \bra{\psi}\mathcal{O}\ket{\psi}
            =
            \int\dd\vf{r}
            \psi^{*}(\vf{r})\mathcal{O}\psi(\vf{r})
            \\
            &=
            \int\dd\vf{r}
            |\psi(\vf{r})|^2
            \frac{\mathcal{O}\psi(\vf{r})}{\psi(\vf{r})},
        \end{align}
        where we have multiplied and divided by the wavefunction. We now define
        the probability density function, $p(\vf{r})$, as
        \begin{align}
            p(\vf{r}) \equiv |\psi(\vf{r})|^2,
        \end{align}
        and the local quantum operator
        \begin{align}
            \mathcal{O}_L(\vf{r})
            &\equiv \frac{\mathcal{O}\psi(\vf{r})}{\psi(\vf{r})}.
        \end{align}
        We see that if the chosen wavefunction is an eigenstate of the quantum
        operator, the local quantum operator becomes a constant.  We can thus
        approximate the expecatation value of the quantum operator using Monte
        Carlo integration.
        \begin{align}
            \expval{O}
            &=
            \int\dd\vf{r}
            p(\vf{r})
            \mathcal{O}_L(\vf{r})
            \approx
            \frac{1}{N}\sum_{i = 1}^N
            p(\vf{r}_i)\mathcal{O}_L(\vf{r}_i).
        \end{align}
        In the case of a non-normalized wavefunction we need to divide by a
        normalization factor in the probability density function.

    \subsection{Local energy}
        As is mostly the case in physics, the quantity we are looking for is the
        energy. We therefore define the \emph{local energy} by
        \begin{align}
            E_L(\vf{r})
            &= \frac{H\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        The energy of the system is thus found by computing the expectation
        value of the local energy.
        \begin{align}
            \expval{E_L}
            =
            \frac{
                \int\dd \vf{r}|\Psi_T(\vf{r})|^2 E_L(\vf{r})
            }{
                \int\dd \vf{r}|\Psi_T(\vf{r})|^2
            }
            \approx
            \frac{1}{N}\sum_{i = 1}^N
            p(\vf{r}_i)E_L(\vf{r}_i).
        \end{align}
        To determine if we have found a minimum we compute the variance of the
        expectation value of the local energy.
        \begin{align}
            \sigma^2 = \langle E_L^2\rangle - \expval{E_L}^2.
        \end{align}
        One of the most computationally intensive parts of the Variational Monte
        Carlo algorithm will be to compute $E_L$. We will therefore find an
        analytical expression for $E_L$ in terms of the trial wavefunctions and
        the system we are exploring.

    \subsection{The drift force}
        A disadvantage in the use of the brute force Metropolis-Hastings
        algorithm is that we might be spending much computational resources in
        an uninteresting part of configuration space. To make smarter moves we
        will add importance sampling (which will be discussed in due time) to
        the algorithm. This alteration of the algorithm needs a new quantity
        called the \emph{drift force} of the system\footnote{Also known as the
        \emph{quantum force}.}.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using the drift force we are able to move towards more interesting parts
        of configuration space as we are following the gradient of the
        wavefunction. We will mainly be interested in the drift force of a
        single particle $k$ as the Monte Carlo sampling involves choosing a
        single particle at a time.

    \subsection{One-Body Densities}
        In order to make some conclusion towards where the particles in our
        systems are most prone to exist, we need a good way to visualise our
        wavefunctions. If we are to visualize the combined wavefunction of two
        three-dimensional particles we need four dimensions per particle and can
        subtract one dimension as we look at a single probability density
        yielding a grand total of seven dimensions. The fact that we will be
        studying systems of up to $N = 500$ particles only adds to the headache.
        We therefore introduce the single-particle density function, or
        \emph{one-body density} $\rho(\vf{r})$. The one-body density is defined
        by the integral over all coordinates except
        one\cite{kristiansen2017time},
        \begin{align}
            \rho(\vf{r}_1)
            \equiv
            \int \dd\vf{r}_2 \dots \dd\vf{r}_N
            \abs{\Psi(\vf{r}_1, \dots, \vf{r}_N)}^2.
        \end{align}
        The one-body density represents the distribution of all the particles in
        the system. Furthermore, $\rho(\vf{r}_1)\dd\vf{r}_1$ will yield the
        probability of finding any particle within the volume element
        $\dd\vf{r}_1$. By convention the one-body density should be normalized
        such that the following relation holds,
        \begin{align}
            N = \int \rho(\vf{r}_1)\dd\vf{r}_1.
        \end{align}
        In other words, the integral over the entire volume in
        $\vf{r}_1$-direction gives us the total number of particles in our
        system. This can be a bit odd at first sight as we are used to getting
        unity when we integrate over the entire probability
        density.\cite{hogberget2013quantum}

        It is best to think of the one-body density as the multivariate
        probability density for the entire system collapsed down to one
        coordinate in a representation than looks very familiar to every
        physicist with a basic knowledge of quantum mechanics.

\section{Systems}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap. The full Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses. The two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction we
    will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        J(\vf{r}),
        %\\
        %&= \Biggl(
        %    \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        %\Biggr)
        %\prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\vf{r} = (\vf{r}_1, \dots, \vf{r}_N)$. The Slater permanent,
    $\Phi_T(\vf{r})$, is given by the $N$ first single-particle functions
    \begin{align}
        \Phi_T(\vf{r})
        &=
        \prod_{i = 1}^N \phi(\vf{r}_i)
        = \prod_{i = 1}^N\exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr].
    \end{align}
    Here $\alpha$ is the variational parameter of the trial
    wavefunction.\footnote{We can also treat $\beta$ as a variational parameter,
    but we will in this project limit ourselves to a single variational
    parameter, i.e., $\alpha$.} The Jastrow factor, $J(\vf{r})$ contributes with
    correlations between the particles and is represented as
    \begin{align}
        J(\vf{r})
        &=
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
    \end{align}
    where the correlation wavefunctions are given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity define the notation
    \begin{gather}
        \phi_i \equiv \phi(\vf{r}_i),
        \\
        r_{jk} \equiv \abs{\vf{r}_j - \vf{r}_k}.
    \end{gather}
    In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.

    \subsection{Non-interacting harmonic oscillators}
        We start by looking at a simple system of non-interacting harmonic
        oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
        hard-shell interaction between the particles and that we use a spherical
        potential trap.  We thus get the trial  wavefunction
        \begin{align}
            \Psi_T(\vf{r})
            = \Phi_T(\vf{r})
            = \prod_{i = 1}^N \exp\para{
                -\alpha |\vf{r}_i|^2
            }.
            \label{eq:trial_simple_gaussian}
        \end{align}
        As $a = 0$ the interaction term, $w(\vf{r}_i, \vf{r}_j)$, vanishes and
        the Hamiltonian is given by (in the spherical case)
        \begin{align}
            H &= \sum_{i = 1}^N h(\vf{r}_i)
            = \sum_{i = 1}^N \Biggl(
                -\frac{\hbar^2}{2m}\nabla_i^2
                + \half m \omega^2 r_i^2
            \Biggr).
        \end{align}
        To find the drift force and the local energy we have to compute the
        gradient and the Laplacian of the trial wavefunction. The gradient is
        given by
        \begin{align}
            \nabla_k\Psi_T(\vf{r})
            &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
        \end{align}
        whereas the Laplacian yields
        \begin{align}
            \nabla^2_k\Psi_T(\vf{r})
            &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}).
            \label{eq:laplacian_simple_gaussian}
        \end{align}
        Here $d$ is the dimensionality of the problem determined by $\vf{r}_k
        \in \mathbb{R}^d$. We can thus use the gradient to find an expression
        for the drift force for particle $k$.
        \begin{align}
            \vf{F}_k(\vf{r})
            &= -4\alpha\vf{r}_k.
        \end{align}
        Using the Laplacian we can compute the kinetic term in the expression
        for the local energy. We get
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{i = 1}^N
            \Biggl(
                -\frac{\hbar^2}{2m}
                \bigl[
                    -2d\alpha + 4\alpha^2 r_i^2
                \bigr]
                + \half m\omega^2 r_i^2
            \Biggr).
        \end{align}
        In natural units, $\hbar = c = 1$, and with the mass set to unity this
        reduces to
        \begin{align}
            E_L(\vf{r})
            &=
            \alpha dN
            + \biggl(
                \half\omega^2
                - 2\alpha^2
            \biggr)
            \sum_{i = 1}^N r_i^2.
            \label{eq:closed_form_natural_units_local_energy}
        \end{align}
        It is worth noting that for $\alpha = \omega/2$\footnote{Note that
        $\alpha$ is required to be positive.} we get a stable value for the
        local energy.
        \begin{align}
            E_L(\vf{r}; \alpha = \omega/2)
            &\equiv
            E_L
            =
            \frac{\omega d N}{2},
            \label{eq:minimum_energy_simple}
        \end{align}
        which turns out to be the exact minimum for the ground state energy of
        this system as we shall see.

        \subsubsection{Exact variational energy}
            As the system is non-interacting and consisting of Gaussians we can
            find an expression for the exact energy as a function of the
            variational parameter $\alpha$, i.e.,
            \begin{align}
                E(\alpha)
                &=
                \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
                =
                \left(
                    \frac{\hbar^2 \alpha}{2m}
                    + \frac{m\omega^2}{8\alpha}
                \right)dN.
                \label{eq:exact_energy}
            \end{align}
            By minimizing this expression with respect to $\alpha$ we get
            \begin{align}
                \dod[]{E(\alpha)}{\alpha} = 0
                \implies
                \alpha_0 = \frac{m\omega}{2\hbar},
            \end{align}
            which in natural units reduces to $\alpha_0 = \omega/2$. The energy
            at this value of $\alpha$ (in natural units) is then
            \begin{align}
                E(\alpha_0)
                &=
                \frac{\omega dN}{2},
            \end{align}
            in perfect accordance with \autoref{eq:minimum_energy_simple} which
            yields high hopes for the latter equation and the expected minimum.

    \subsection{Interacting hard sphere bosons}
        Moving to the full elliptical system for $\beta \neq 1$ and setting $a
        \neq 0$ we use the trial wavefunction from
        \autoref{eq:initial_trial_wavefunction}. We rewrite the Jastrow factor
        to
        \begin{align}
            J(\vf{r})
            &=
            \exp\para{
                \sum_{j < l}^N u(r_{jl})
            },
        \end{align}
        where
        \begin{align}
            u(r_{jk})
            = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr]
            \equiv u_{jk}
        \end{align}
        The drift force for particle $k$ in this system becomes
        \begin{align}
            \vf{F}_k(\vf{r})
            &=
            2\Biggl(
                \frac{\nabla_k\phi_k}{\phi_k}
                + \sum_{m \neq k}^N
                \nabla_k u_{km}
            \Biggr).
        \end{align}
        The Laplacian of particle $k$ divided by the trial wavefunction is then
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and the Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            =
            -2\alpha
            (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k),
            \\
            \frac{\nabla_k^2\phi_k}{\phi_k}
            =
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr).
        \end{gather}
        Repeating this for the correlation functions yields
        \begin{align}
            \dpd[]{u_{km}}{r_{km}}
            &=
            \frac{a}{r_{km}(r_{km} - a)},
            \\
            \dpd[2]{u_{km}}{r_{km}}
            &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
        \end{align}
        The local energy is thus
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                + v(\vf{r}_k)
            },
        \end{align}
        where the Laplacian is given by \autoref{eq:laplacian_full_system}.  We
        will avoid writing out the full expression for the local energy for this
        system due to the size of the equation.

        \subsubsection{Scaling the system}
            We now introduce a scaled distance $\vf{r}' = \vf{r}/a_{\text{ho}}$
            \cite{dubois2001bose}, where
            \begin{align}
                a_{\text{ho}}
                &=
                \sqrt{\frac{\hbar}{m\omega}}.
            \end{align}
            We can then rewrite the Hamiltonian for the elliptic potential in terms
            of this new distance. By doing a variable substitution for each
            direction in the Laplace operator we get
            \begin{align}
                \nabla_k^2
                &=
                \frac{1}{a_{\text{ho}}^2}{\nabla_k'}^2.
            \end{align}
            The one-body part of the Hamiltonian then yields
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The wavefunction does not need to be scaled as the
            exponentials are already dimensionless. We only need to use the
            scaled hard core radius $a' = a/a_{\text{ho}}$. We are interested in
            using the radius $a' = 0.0043$ as in the references
            \cite{dubois2001bose} and \cite{nilsen2005vortices}.

\section{Algorithms}
    In this project we rely on a Monte Carlo approach of random sampling to
    obtain numerical results.  We simulate random walks over a volume in order
    to find optimal parameters in our trial wavefunctions.  The most common of
    such methods, which we make use of herein, is the Metropolis-Hastings
    algorithm.

    \subsection{The Metropolis-Hastings Algorithm}
        The Metropolis-Hastings algorithm can in our particular situation be
        condensed down to the following steps:

        \begin{enumerate}
            \item The system is initialised by a certain number $N$ of randomly
                generated positions, $\vf{r} = (\vf{r}_1, \dots, \vf{r}_N)$,
                which represent the particles\footnote{Also known as
                \emph{random walkers}.}. This allows us to evaluate the
                wavefunction at these points and compute the local energy
                $E_L(\vf{r})$.

            \item The initial configuration is changed by setting a new position
                for one of these particles. The particle is picked at random and
                is only allowed to move a predetermined step length.

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                transition probability decides if the particle move is rejected
                or accepted.

            \item If the particle movement is accepted we compute the local
                energy for the new system. If not, we reset the system to the
                previous state and reuse the last local energy.

            \item Accumulate the local energy. Repeat the steps until
                convergence is reached or a set number of iterations has been
                done. When finished divide the accumulated energy by the number
                of cycles to compute the expectation value of the energy.

        \end{enumerate}

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a \emph{transition
        probability}. The strength of the algorithm lies in that the transition
        algorithm need not be known. For example, the simplest case is to accept
        the new state, i.e., the new position for the random walker, if the
        ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$. This transition criteria
        results in what we call the brute force Metropolis-Hastings algorithm.
        By counting the number of accepted steps and dividing by the total
        number of Monte Carlo cycles we get the \emph{acceptance probability},
        denoted $A$.  The acceptance probability helps in choosing a decent
        length for the proposed new step. If the acceptance probability is very
        low we can lower the step length to get more accepted steps and vice
        versa.

        \subsubsection{Initial distribution of particles}
            Prior to performing the Metropolis-Hastings algorithm we have to
            have an initial distribution of the particles. As the algorithm is
            supposed to move particles in the right direction this distribution
            can be \emph{somewhat} arbitrary. There are however a few caveats to
            this approach.

            \begin{enumerate}
                \item The distribution is unphysical. This results in the
                    particles taking up positions which can yield bogus values
                    for the local energy. A remedy is to allow for some
                    \emph{thermalization steps}, i.e., scatter the particles and
                    run the algorithm for a set number of steps prior to
                    starting the actual sampling.

                \item The initial \emph{spread} of the particles is too small or
                    too large. The former can lead to an acceptance ratio of
                    zero if no steps are accepted due to overlap. This is
                    something that can happen to the interacting system we are
                    studying as an overlap between the particles makes the
                    wavefunction evaluate to zero thus rejecting all steps in
                    the ratio test.  The latter can lead to the system never
                    reaching a true state as they are scattered too far away
                    from the interesting regions in the configuration space.

                \item The system imposes some restrictions on the distribution.
                    This can mean that we have to distribute the particles in a
                    certain way to make sure that we don't break the
                    restrictions. This resembles the former case when the spread
                    is too small for our systems.
            \end{enumerate}

            In the case of the interacting system we are looking at we have to
            take some measures to make sure that the initial spread does not
            break the Monte Carlo cycles. We therefore distribute all the
            particles on a grid with a distance larger than the diameter of the
            hard sphere before letting the system thermalize. After the
            thermaliztion we start the sampling.

        \subsubsection{Importance Sampling}
            A problem with the naïve Metropolis-Hastings sampling approach
            described above is that the sampling of position space is done with
            no regard for where we are likely to find a particle. This problem
            can be remedied through \emph{importance sampling}.  It is
            reasonable to assume that the particles we erratically scatter in
            space are prone to movement towards the peaks of the probability
            density as dictated by the wave function. Consider therefore the
            Fokker-Planck equation,
            \begin{align}
                \dpd[]{\Psi_T}{t}
                &=
                D \nabla\cdot
                \left(
                    \nabla
                    - \vf{F}
                \right) \Psi_T,
                \label{eq:fokker_planck}
            \end{align}
            which describes the evolution in time of a probability density
            function. In our case this is the trial wavefunction
            $\Psi_T$\footnote{The trial wavefunction is not a probability
            density, but the Fokker-Planck equation is robust enough to handle
            the wavefunction as is.}.  Originaly an equation that models
            diffusion, we have a diffusion term $D$ and the drift force
            \autoref{eq:drift_force}. In our case the diffusion term $D$ is
            simply $1/2$ from the kinetic energy (in natural units).

            We use the Langevin equation to find the new position of the
            particle.
            \begin{align}
                \dpd[]{\vf{r}}{t}
                &=
                D\vf{F}(\vf{r}) + \vfg{\eta},
                \label{eq:langevin}
            \end{align}
            where $\vfg{\eta}$ is a uniformly distributed stochastic variable
            for each dimenion.  Solving Langevin's equation by Euler's method
            gives a recursive relation for the subsequent new positions of a
            particle.
            \begin{align}
                \vf{r}_{i + 1}
                &=
                \vf{r}_i + D\vf{F}(\vf{r}_i)\delta t
                + \vfg{\xi}\sqrt{\delta t},
            \end{align}
            for a given time step $\delta t$
            \footnote{
                Bear in mind that \autoref{eq:langevin} is only valid as $\Delta
                t \to 0$, a property stemming from the use of Euler's method.
            }
            and a normally distributed stochastic variable $\vfg{\xi}$.

            Now we need to change the ratio computation in the
            Metropolis-Hastings algorithm to something that takes the new
            sampling method into account.
            \begin{align}
                q(\vf{r}_{i + 1}, \vf{r}_i)
                &=
                \frac{
                    G(\vf{r}_{i}, \vf{r}_{i + 1}, \delta t)
                    \left|\Psi_T(\vf{r}_{i + 1})\right|^2
                }
                {
                    G(\vf{r}_{i + 1}, \vf{r}_{i}, \delta t)
                    \left|\Psi_T(\vf{r}_{i})\right|^2
                },
            \end{align}
            where $G$ is the Green's function of the Fokker-Planck equation
            given by
            \begin{align}
                G(\vf{r}_{i}, \vf{r}_{i + 1}, \delta t)
                &=
                \exp\Biggl(
                    -\frac{\bigl[
                        \vf{r}_{i}
                        - \vf{r}_{i + 1}
                        - D\vf{F}(\vf{r}_{i + 1})\delta t
                    \bigr]^2}{4D\delta t}
                \Biggr)
                \nonumber \\
                &\qquad
                \times
                \frac{1}{(4\pi D\delta t)^{dN/2}},
            \end{align}
            where $d$ is the dimensionality. The new ratio, $q$, is still
            subjected to the same ratio test as in the brute force
            Metropolis-Hastings algortihm.

    \subsection{Statistical Analysis}
        If the results of the metropolis sampling were completely uncorrelated,
        it would be enough to compute the standard deviation in a familiar way,
        \begin{equation}
            \sigma = \sqrt{\frac{1}{N}(\ev{E^2_L} - \ev{E_L}^2},
            \label{eq:standard_deviation}
        \end{equation}
        where $N$ is the number of samples, or Monte-Carlo cycles, in the
        experiment.  However, it is reasonable to assume that the data we are
        dealing with in this study is liable to suffer from
        \emph{autocorrelation} and \autoref{eq:standard_deviation} does not
        hold. The prevailing definition of autocorrelation in a data stream or
        signal is correlation between a delay of the signal and the original
        signal. One would be interested to find the delay, or lag, in the signal
        at which the "self-correlation" is highest. We refer to this spacing as
        $d$, and define the following correlation function,
        \begin{equation}
            f_d
            = \frac{1}{n -d} \sum_{k=1}^{n-d}
            (x_k - \bar{x}_n)(x_{k+d} - \bar{x}_n).
            \label{eq:correlation_f_d}
        \end{equation}
        The keen reader would have noticed that the function $f_d$ in
        \autoref{eq:correlation_f_d} would be equal to the sample variance for
        $d=0$. We can now define the \emph{autocorrelation function}
        \begin{equation}
            \kappa_d = \frac{f_d}{\text{Var}(x)},
            \label{eq:autocorrelation_function}
        \end{equation}
        which is equal to $1$ if the data exhibits no autocorrealtions, equating
        to $d=0$. From the autocorrelation function in
        \autoref{eq:autocorrelation_function} we in turn define the
        \emph{autocorrelation time},
        \begin{equation}
            \tau = 1 + 2 \sum_{d=1}^{n-1}\kappa_d.
        \end{equation}
        Notice that the autocorrelation time is $1$ for a correlation free
        experiment.

        We are now able to make a correction to the expression for the standard
        deviation improving on \autoref{eq:standard_deviation} by taking
        correlation into account,
        \begin{equation}
            \sigma =\sqrt{\frac{1 + 2\tau / \Delta t}{N}
            (\ev{E^2_L} - \ev{E_L}^2)},
        \end{equation}
        where $\Delta t$ is the time between each sample. The main problem at
        this point is to find $\tau$, which we do not know for any given system
        and it is generally very expensive to compute. In order to find a good
        estimate of $\tau$ we use a procedure called \emph{blocking}.

        \subsubsection{Blocking}
            In the method of blocking we group the samples into blocks of
            increasing size. If one where to compute the standard deviation for
            each block, one should see the variance increase with the block
            size.  The standard deviation would only increase up to a certain
            point, from whence it would stay almost constant. What is happening
            is that we have reached a point where a particular sample from one
            block is no longer correlated with a corresponding sample from an
            adjacent block.  The block size for this point of convergence now
            functions as an estimate for the autocorrelation time $\tau$.

            Instead of going by this "manual" method of looking at charts, we
            will construct a test statistic and let the computer do the
            necessary considerations in a more "automatic way.

>>>>

\title{Ground state energy of quantum dots using the coupled cluster method}
\author{Winther-Larsen, Sebastian Gregorius}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\author{Schøyen, Øyvind Sigmundson}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\date{\today}

\begin{abstract}
    Something about coupled-cluster... Preferably doubles.
\end{abstract}

@@

\bibliography{references}
